Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-26 review/Response to reviews.Rmd

While we could not test for an interaction between habitat and lagged SPEI (i.e. whether the shape of the crossbasis smooth was *significantly* different between habitats), we did test for a main effect of habitat by looking for overlap in 84% confidence intervals of model intercepts (see also response to \@ref(R2-6)).
However, to reduce confusion here, we have changed "significantly" to "substantially" so as not to imply statistical significance for all of "magnitude, direction and lag time".

##  {#R2-11}

> Line 370: There are no confidence intervals in Fig 2, so we don't know if the difference is significant.

Figure 2 shows the survivorship curve of plants tagged in 1998.
These are exact counts and therefore do not have confidence intervals.
They do not require statistics to interpret if they are different.
We think the confusion may have come from the axis scale and label ("P(survived)") which we intended to be interpreted as "proportion of plants surviving" but may have been interpreted as "probability of survival". We have changed the scale to a percentage and the axis label to "% surviving".

### 2021-08-26 review/Response to reviews.Rmd

> One challenge of using a new modeling approach is making the methods of the paper understandable and to stand alone.
> As written in the methods and results, I found the implementation very difficult to follow.
> A slightly longer summary of the methods would help.

We have added an additional paragraph to the methods describing how the crossbasis smooth was fit using the `mgcv` and `dlnm` packages.
<!--# add line reference before submiting! --> We are not entirely sure if this is the type of description the reviewer was looking for.
If the reviewer(s) or editor see this as too detailed for the main text, we are happy to remove it, move it to a supplement, or simply replace it with a reference to the (now available) archived code for the analysis.
See also our response to \@ref(R3-3).

> In the results, I was surprised to see only a few p-values and not much else about the models in terms of how they were assessed for goodness of fit.
> While I am not at all suggesting that p-values are the solution, I would have liked a little more sense of how well the models fit.
> One way to do that might be to show the raw data plotted with the model fits so the reader can see, for example, what kind of functions were drawn.
> Another might be to present something in table format (the comparable version of a table of AICs or of the results of frequentist tests).

Because model selection was performed automatically by shrinkage using the `select = TRUE` option in the `gam()` function, we cannot show relative model fits (i.e., we did not perform model selection and can't show dAIC).
However, we have now included a table with R^2^ values for each model indicating fit as well as estimated degrees of freedom and p-values for each term.
We have replaced many of the parenthetical p-values reported in the results section with references to this table.

@@

Figure 2 shows the survivorship curve of plants tagged in 1998.
These are exact counts and therefore do not have confidence intervals.
They do not require statistics to interpret if they are different.
We think the confusion may have come from the axis scale and label ("P(survived)") which we intended to be interpreted as "proportion of plants surviving" but may have been interpreted as "probability of survival". We have changed the scale to a percentage and the axis label to "% surviving".

##  {#R2-12}

> Lines 383 ff: see comment 7

<!--# although we didn't test for a significant interaction, they have different areas of significance and different ranges of fitted values. Not sure what else to say about this-->

@@

...see also our response to \@ref(R1-5).
<!--# R1-5 is asking for more discussion on limitations of DLNMs -->

##  {#R2-14}

> 424-427: I think this is exaggerating since the differences between habitat types is not significant.

<!--# the magnitude *was* greater, interaction not tested -->...see also our response to \@ref(R2-10)

### 2021-08-26 doc/paper.Rmd
add results table.  closes #78

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.93$ in both continuous forest and forest fragments).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e. negative growth in Figure \@ref(fig:eda)a).
When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; Table \@ref(tab:intercept-table)).
However, survival in both habitats was size dependent ($p < 0.001$ for the effect of log-transformed plant size in year *t* on survival in year *t+1* in both habitats).
The survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)b), but the smallest plants had higher survival in 1-ha fragments.

@@

*Flowering:* The overall proportion of plants that flowered was very low.
Flowering probability was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively); this difference was statistically significant (Table \@ref(tab:intercept-table)).
The observed disparity was largely due to the fact that flowering is also size-dependent ($p < 0.001$ in both habitats), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Despite the flowering probability of the largest plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. 17% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

Drought history had a significant ($p < 0.001$) effect on the survival, growth, and flowering of plants in both habitats.
Comparing the respective crossbasis surfaces, however, reveals that the specific climatic drivers, their timing, and their impact on individual vital rates all differed among habitats.

@@

*Growth:* The effects of drought history on trends in plant size were generally similar for continuous forest and fragments.
However, the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)).
For example, under average conditions (i.e., SPEI = 0), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
However, the growth of plants in fragments is reduced when the current wet season is unusually wet (i.e., SPEI\>2), and as a result the plants in continuous forests will be larger by up to log(size) = 0.57.
In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, with a more pronounced response in 1-ha fragments.

*Flowering:* Overall, the the probability of flowering was higher in continuous forest than in 1-ha fragments for all values of SPEI (Figure \@ref(fig:flwr-spei)), although this difference was not significant (84% CIs of intercepts overlap).
SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
This led to some important inter-habitat differences in plant responses to prior droughts.
In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20) and in the wet season 34--36 months prior all increased the probability of flowering.
However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
In fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger (CF: $p = 0.048$; 1-ha: $p = 0.030$) and flower again (CF: $p < 0.001$; 1-ha: $p = 0.004$).
The random effect of plant ID on flowering probability was also significant (CF: $p < 0.001$; 1-ha: $p=0.003$), indicating significant individual-level variation in flowering probability.

Finally, with the exception of survival in 1-ha fragments ($p = 0.253$), the delayed effects of SPEI on all three vital rates varied significantly among plots ($p < 0.01$ for the random effect of plot).

@@

```{r tab.cap=ci_cap, label="intercept-table"}
tar_read(intercept_table)
```

>>>>

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year ($P_{surv} = 0.93$ in both continuous forest and forest fragments).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e. negative growth in Figure \@ref(fig:eda)a).
When summarizing across years, plots, and plant sizes, the survival probability of *Heliconia acuminata* was similarly high in both continuous forest and fragments ($P_{surv}=0.95$; Figure \@ref(fig:eda)b; Table \@ref(tab:intercept-table)).
However, survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
The survival probability of large plants approached 1 in both habitat types, but the smallest plants had higher survival in 1-ha fragments.

@@

*Flowering:* The overall proportion of plants that flowered was very low.
Flowering probability was 31% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively); this difference was statistically significant (Table \@ref(tab:intercept-table)).
The observed disparity was largely due to the fact that flowering is also significantly size-dependent (Table \@ref(tab:results-table)), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 (i.e., log(size) \> 5 in Figure \@ref(fig:size-plot)c).
Despite the flowering probability of the largest plants being greater in fragments than continuous forest, populations in fragments had proportionately fewer plants above the reproductive size threshold (Figure \@ref(fig:size-plot)d).
The most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering reproductive-sized plants was 28% in continuous forest vs. 17% in 1-ha fragments (Figure \@ref(fig:eda)c).

@@

Drought history had a significant effect on the survival, growth, and flowering of plants in both habitats (Table \@ref(tab:results-table)).
Comparing the respective crossbasis surfaces, however, reveals that the specific climatic drivers, their timing, and their impact on individual vital rates all differed among habitats.

@@

*Growth:* The effects of drought history on trends in plant size were generally similar for continuous forest and fragments.
However, the crossbasis function for 1-ha fragments indicated more complex responses in some situations (edf = 17.8 for 1-ha fragments; edf = 13.0 for continuous forest; see also Figure \@ref(fig:growth-spei)) <!--# this is no longer true.  need to update -->.
For example, under average conditions (i.e., SPEI = 0), growth is similar or slightly higher in continuous forest over all lag periods (Figure \@ref(fig:growth-spei)c).
However, the growth of plants in fragments is reduced when the current wet season is unusually wet (i.e., SPEI\>2), and as a result the plants in continuous forests will be larger by up to log(size) = 0.57.
In contrast, drought at lags of 8--11 months (i.e., the end of the preceding year's wet season) led to increased growth in both habitats, with a more pronounced response in 1-ha fragments.

*Flowering:* Overall, the the probability of flowering was significantly higher in continuous forest than in 1-ha fragments (Figure \@ref(fig:flwr-spei), Table \@ref(tab:intercept-table)).
SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure \@ref(fig:flwr-spei)b).
This led to some important inter-habitat differences in plant responses to prior droughts.
In continuous forests, recent drought (i.e., at lag = 0--2 with SPEI \< -1), drought two dry seasons prior (lags 15--20) and in the wet season 34--36 months prior all increased the probability of flowering.
However, drought at the end of the rainy season one year prior (lags 7--13) significantly reduced flowering probability (Figure \@ref(fig:flwr-spei)a).
In fragments, recent drought had no significant effect on flowering probability and only drought at two dry seasons prior (lags 16--20) and in the wet season 33--36 months prior increased flowering probability (Figure \@ref(fig:flwr-spei)b).
The effects of drought on flowering probability were strong in continuous forest compared to 1 ha fragments (Figure \@ref(fig:flwr-spei)c).
We found no evidence for a cost of reproduction: in both forest and fragments, plants that had flowered in the previous year were on average more likely to be larger and flower again (Table \@ref(tab:results-table)).
The random effect of plant ID on flowering probability was also significant (Table \@ref(tab:results-table)), indicating significant individual-level variation in flowering probability.
<!--# individual level random effects need to be removed? -->

Finally, with the exception of the model for survival in 1-ha fragments, the random effect of plot was significant, indicating that vital rates varied among plots (Table \@ref(tab:results-table)).

@@

```{r label="intercept-table"}
tar_read(intercept_table) %>% set_caption(ci_cap, style = "Image Caption")
```

<br>

```{r}
result_cap <- 'Marginal hypothesis tests for models of survival probability, size, and flowering probability of *H. acuminata* in continuous forest and 1 ha fragment plots. The adjusted R^2^ is reported as a measure of model fit. The model terms included the parametric fixed effect factor of whether plants flowered the previous year ("flowered"), the smoothed fixed effect of plant size ("s(log(size~t~))"), the crossbasis smooth of lagged SPEI ("s(SPEI, lag)"), and a random effect of plot ID, ("s(plot)"). Degrees of freedom are reported for the parametric term, and estimated degrees of freedom (edf) are reproted for smooths.  Larger values for edf indicate more complex smooths and when edf is zero the term is effectively dropped from the model. The test statistic reported is $\\chi^2$ for survival and flowering and $F$ for log(size~t+1~).'
```

<!--# add something about how p-values are appropriate for random effects?  see help files from mgcv for references -->

```{r label="results-table"}
tar_read(results_table) %>% set_caption(result_cap, style = "Image Caption")
```
